A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2015-09-03 docs/discussion_ML_v1.01.tex
Finished section on JAM theory.

Further knowledge of the stellar populations within J1331 and their initial mass functions (IMF) would be useful. In particular a sophisticated guess for the \emph{stellar} mass-to-light ratio in J1331's bulge could be compared to our very reliable measurement of the \emph{total} mass-to-light ratio inside the Einstein radius (see Table [TO DO: REF]) to support or contradict the presence of a lot of dark matter in the bulge. Traditional choices for the IMF are the bottom-heavy IMF by \citet{Salpeter1955}
$$\xi(m) \propto m^{-x}, x=2.35$$
and the IMFs by \citet{2002Sci...295...82K} and \citet{Chabrier2003}, which are in agreement with each other and predict less low-mass stars. \\\citet{Ferreras} found a relation between the central stellar velocity dispersion $\sigma_0$ in early-type galaxies and the IMF slope. A higher $\sigma_0$ suggests a more bottom-heavy IMF. For an unimodal (Salpeter-like) IMF and a $\sigma_0 \propto 200 $ km/s for J1331 (see Figure \ref{fig:kinematics}) they predict $x \approx 2.33$ (see Figure 4 in \citet{Ferreras}), which is close to the standard Salpeter slope. These findings are supported by \citet{2014MNRAS.438.1483S}, who found for SDSS galaxies also a relation between unimodal (single power-law) IMF slope and stellar velocity dispersion. Their Table 2 lists $x(\sigma_*=190 \pm 20 \text{ km s}^{-1}) = 2.08 \pm 2.08$ and $x(\sigma_*= 230 \pm 20 \text{ km s}^{-1}) = 2.33 \pm 0.4$. When assuming a bi-modal (Kroupa-equivalent-like) IMF, \citet{Ferreras} predict $x \approx 2.85$ (see Figure 4 in \citet{Ferreras}) for J1331's central velocity dispersion. This is more bottom-heavy than the standard \citet{2002Sci...295...82K} IMF. Overall the central velocity dispersion suggests a rather bottom-heavy IMF J1331's bulge and therefore large stellar mass-to-light ratio.

@@

\\We compare our findings of the stellar mass-to-light ratios with the study by \cite{SWELLSV}: \cite{SWELLSV} found that the bulge of J1331 has an IMF more bottom-heavy than the Salpeter IMF from fitting a NFW halo to (1) the Einstein mass and (2) gas kinematics at larger radii $\gtrsim 8$ arcsec. We fitted the stellar kinematics within $\sim 3.5$ arcsec and the Einstein mass with a NFW halo and found $\Upsilon^\text{dyn}_{*,I} = 4.2 \pm 0.2$ (see Table [TO DO ???]). \cite{SWELLSV}  found systematically lower NFW halos ($v_\text{circ}(5\text{arcsec}) \sim 120 \text{km s}^{-1}$ according to their Figure 2) than we did ($v_\text{circ}(5\text{arcsec}) \sim 200 \text{km s}^{-1}$, see Figure [TO DO ???]). [TO DO: Continue writing notes into text]

@@

\item We found an Einstein mass-to-light ratio of $\Upsilon_\text{I,tot,Ein} = 5.6$ (see Table [TO DO ???]). In case of a lower mass NFW halo, the largest contribution to this mass-to-light ratio would then be stellar mass. And the corresponding IMF would be more bottom heavy than Salpeter as well, like what \citet{SWELLSV} found.
\item The light distribution multiplied with $\Upsilon_\text{I,tot,Ein}$ looks very similar to \cite{SWELLSV} [TO DO: Look details up in their paper and my work

@@

\begin{table}
\centering
\begin{tabular}{cccccc}
\hline\hline
& & \multicolumn{2}{c}{Chabrier IMF} & \multicolumn{2}{c}{Salpeter IMF}\\
      &  $L$ [$10^{10}L_{\odot}$]                & $M_*$ [$10^{10}M_\odot$]               & $\Upsilon_*^\text{chab}$ & $M_*$ [$10^{10}M_\odot$] & $\Upsilon_*^\text{sal}$ \\\hline
bulge &   $3.10 \pm 0.15 $  & $7.8 \pm 1.8$ & $2.5 \pm 0.6$ & $14.5 \pm 3.7 $ & $4.7 \pm 1.2$ \\
disk  &   $2.35 \pm 0.11 $  & $2.9 \pm 0.7$ & $1.2 \pm 0.3$ & $5.2 \pm 1.1$ & $2.2 \pm 0.5$ \\
total &   $5.45 \pm 0.19$ & $10.6 \pm 1.9$& & $19.7 \pm 3.9$&\\\hline
\end{tabular}
\caption{Total I-band luminosity, stellar mass and mass-to-light ratio, calculated from the I-band AB magnitudes and stellar masses found for J133's bulge and disk by \citet{SWELLS} (their table 2) for comparison with this work. The transformation from AB magnitudes to the Johnson-Cousins I-Band used the relation $I[\text{mag}] = I[\text{ABmag}] - 0.309$ from [TO DO: REF: Frei and Gunn 1994] (their table 2). For the conversion from apparent magnitude to total luminosity the redshift $z=0.113$ [TO DO: REF: Brewer] was turned into a luminosity distance using the cosmology by [TO DO: REF: Dunkley]. }
\label{tab:previousresults}
\end{table}

>>>>

Further knowledge of the stellar populations within J1331 and their initial mass functions (IMF) would be useful. In particular a sophisticated guess for the \emph{stellar} mass-to-light ratio in J1331's bulge could be compared to our very reliable measurement of the \emph{total} mass-to-light ratio inside the Einstein radius (see Table \ref{tab:bestfitlensmodel}) to support or contradict the presence of a lot of dark matter in the bulge. Traditional choices for the IMF are the bottom-heavy IMF by \citet{Salpeter1955},
$$\xi(m) \propto m^{-x}, x=2.35,$$
(where $\xi(m) \diff m$ is the number of stars with mass $m$ in $[m,m+\diff m]$) and the IMFs by \citet{2002Sci...295...82K} and \citet{Chabrier2003}, which are in agreement with each other and predict less low-mass stars. \\\citet{Ferreras} found a relation between the central stellar velocity dispersion $\sigma_0$ in early-type galaxies and the IMF slope. A higher $\sigma_0$ suggests a more bottom-heavy IMF. For an unimodal (Salpeter-like) IMF and a $\sigma_0 \propto 200 $ km/s for J1331 (see Figure \ref{fig:kinematics}) they predict $x \approx 2.33$ (see Figure 4 in \citet{Ferreras}), which is close to the standard Salpeter slope. These findings are supported by \citet{2014MNRAS.438.1483S}, who found for SDSS galaxies also a relation between unimodal (single power-law) IMF slope and stellar velocity dispersion. Their Table 2 lists $x(\sigma_*=190 \pm 20 \text{ km s}^{-1}) = 2.08 \pm 2.08$ and $x(\sigma_*= 230 \pm 20 \text{ km s}^{-1}) = 2.33 \pm 0.4$. When assuming a bi-modal (Kroupa-equivalent-like) IMF, \citet{Ferreras} predict $x \approx 2.85$ (see Figure 4 in \citet{Ferreras}) for J1331's central velocity dispersion. This is more bottom-heavy than the standard \citet{2002Sci...295...82K} IMF. Overall the central velocity dispersion suggests a rather bottom-heavy IMF J1331's bulge and therefore large stellar mass-to-light ratio.

@@

\\We compare our findings of the stellar mass-to-light ratios with the study by \cite{SWELLSV}: \cite{SWELLSV} found that the bulge of J1331 has an IMF more bottom-heavy than the Salpeter IMF from fitting a NFW halo to (1) the Einstein mass and (2) gas kinematics at larger radii $\gtrsim 8$ arcsec. We fitted the stellar kinematics within $\sim 3.5$ arcsec and the Einstein mass with a NFW halo and found $\Upsilon^\text{dyn}_{*,I} = 4.2 \pm 0.2$ (see Table \ref{tab:modelB4_bestfit}). \cite{SWELLSV}  found systematically lower NFW halos ($v_\text{circ}(5\text{arcsec}) \sim 120 \text{km s}^{-1}$ according to their Figure 2) than we did ($v_\text{circ}(5\text{arcsec}) \sim 200 \text{km s}^{-1}$, see Figure \ref{fig:kinematics}). [TO DO: Continue writing notes into text]

@@

\item We found an Einstein mass-to-light ratio of $\Upsilon_\text{I,tot,Ein} = 5.6$ (see Table \ref{tab:bestfitlensmodel}). In case of a lower mass NFW halo, the largest contribution to this mass-to-light ratio would then be stellar mass. And the corresponding IMF would be more bottom heavy than Salpeter as well, like what \citet{SWELLSV} found.
\item The light distribution multiplied with $\Upsilon_\text{I,tot,Ein}$ looks very similar to \cite{SWELLSV} [TO DO: Look details up in their paper and my work]

@@

\begin{table}
\centering
\begin{tabular}{cccccc}
\hline\hline
& & \multicolumn{2}{c}{Chabrier IMF} & \multicolumn{2}{c}{Salpeter IMF}\\
      &  $L$ [$10^{10}L_{\odot}$]                & $M_*$ [$10^{10}M_\odot$]               & $\Upsilon_*^\text{chab}$ & $M_*$ [$10^{10}M_\odot$] & $\Upsilon_*^\text{sal}$ \\\hline
bulge &   $3.10 \pm 0.15 $  & $7.8 \pm 1.8$ & $2.5 \pm 0.6$ & $14.5 \pm 3.7 $ & $4.7 \pm 1.2$ \\
disk  &   $2.35 \pm 0.11 $  & $2.9 \pm 0.7$ & $1.2 \pm 0.3$ & $5.2 \pm 1.1$ & $2.2 \pm 0.5$ \\
total &   $5.45 \pm 0.19$ & $10.6 \pm 1.9$& & $19.7 \pm 3.9$&\\\hline
\end{tabular}
\caption{Total I-band luminosity, stellar mass and mass-to-light ratio, calculated from the I-band AB magnitudes and stellar masses found for J133's bulge and disk by \citet{SWELLS} (their table 2) for comparison with this work. The transformation from AB magnitudes to the Johnson-Cousins I-Band used the relation $I[\text{mag}] = I[\text{ABmag}] - 0.309$ from \citet{FG1994} (their table 2). For the conversion from apparent magnitude to total luminosity the redshift $z=0.113$ \citet{SWELLSIII} was turned into a luminosity distance using the cosmology by \ref{WMAP5cosm}. }
\label{tab:previousresults}
\end{table}

### 2015-09-03 docs/model_JAM_v1.01.tex
Finished section on JAM theory.

Jeans axisymmetric models (JAM) assume galaxies (a) to be collisionless, i.e. the collisionless Boltzmann equation for the distribution function $f(\vect{x},\vect{v},t)$ has to be satisfied ($\frac{\diff f(\vect{x},\vect{v},t}{\diff t} = 0$), (b) in a steady state ($\frac{\partial}{\partial t} = 0$), (c) axisymmetric (best described in cylindrical coordinates $(R,z,\phi)$ and $\frac{\partial}{\partial \phi} = 0$). From this follow the axisymmetric Jeans equations as the vector-valued first moment of the Boltzmann equation, i.e.
\begin{equation*}
\int \frac{\diff f}{\diff t} \Diff3 v = 0.
\end{equation*}
To be able to solve the Jeans equations, additional assumptions about the velocity ellipsoid tensor $\langle v_i v_j\rangle$ have to be made. We follow \citet{Cap08} and assume firstly, that the galaxy's velocity ellipsoid is aligned with the cylindrical coordinate system, i.e. $\langle v_i v_j\rangle = 0$ for $i\neq j$. Secondly, we assume a constant ratio between the radial and vertical 2nd velocity moments, $\beta_z \equiv 1 - \langle v_z^2 \rangle / \langle v_R^2\rangle$. This reduces the Jeans equations to two equations for $\langle v_z^2 \rangle$ and $\langle v_\phi^2 \rangle$, that can be solved by means of one integration,
\begin{eqnarray*}
n \langle v_z^2 \rangle (R,z) &=& \int_0^\infty n \frac{\partial \Phi}{\partial z} \diff z\\
n \langle v_\phi^2 \rangle (R,z) &=& R \frac{\partial}{\partial R} \left( \frac{n \langle v_z^2 \rangle}{1-\beta_z} \right) + \frac{n \langle v_z^2 \rangle}{1 - \beta_z} + R n \frac{\partial \Phi}{\partial R},
\end{eqnarray*}
where $n(\vect{x}) = \int f(\vect{x},\vect{v}) \diff3 v$ is the number density of tracers and $\Phi(\vect{x})$ the galaxy's gravitational potential, generated by the mass density $\rho(\vect{x})$ via Poisson's equation.
\\The JAM modelling approach by \citet{Cap08} makes use of expressing the tracer density and the mass density as MGEs (see also \citet{1994A&A...285..723E}). The density of stellar tracers is assumed to be proportional to the observed and deprojected brightness distribution $\nu(R,z)$ in Eq. (\ref{eq:deprojMGE}). The mass density consists of several sets of MGEs: One MGE, that is usually taken to be $\nu(R,z)$ multiplied by a constant stellar mass-to-light ratio $\Upsilon_*$, describes the distribution of stellar mass in the galaxy. To mimick gradients of stellar mass-to-light ratio, each Gaussian could be assigned its own $\Upsilon_{*,i}$. To add a Navarro-Frenck-White (NFW) \citep{Navarro+1995c,NFW96} dark matter halo component, a MGE generated from a fit to a NFW profile can be added to the stellar component. [TO DO: continue on p. 74]

>>>>

Jeans axisymmetric models (JAM) assume galaxies (a) to be collisionless, i.e. the collisionless Boltzmann equation for the distribution function $f(\vect{x},\vect{v},t)$ has to be satisfied ($\frac{\diff f(\vect{x},\vect{v},t}{\diff t} = 0$), (b) in a steady state ($\frac{\partial}{\partial t} = 0$), (c) axisymmetric (best described in cylindrical coordinates $(R,z,\phi)$ and $\frac{\partial}{\partial \phi} = 0$). From this follow the axisymmetric Jeans equations as the vector-valued first moment of the Boltzmann equation, i.e.
\begin{equation*}
\int \frac{\diff f}{\diff t} \Diff3 v = 0.
\end{equation*}
To be able to solve the Jeans equations, additional assumptions about the velocity ellipsoid tensor $\langle v_i v_j\rangle$ have to be made. We follow \citet{Cap08} and assume firstly, that the galaxy's velocity ellipsoid is aligned with the cylindrical coordinate system, i.e. $\langle v_i v_j\rangle = 0$ for $i\neq j$. Secondly, we assume a constant ratio between the radial and vertical 2nd velocity moments, 
\begin{equation}
\beta_z \equiv 1 - \langle v_z^2 \rangle / \langle v_R^2\rangle. \label{eq:bz}
\end{equation}
This reduces the Jeans equations to two equations for $\langle v_z^2 \rangle$ and $\langle v_\phi^2 \rangle$, that can be solved by means of one integration,
\begin{eqnarray}
n \langle v_z^2 \rangle (R,z) &=& \int_0^\infty n \frac{\partial \Phi}{\partial z} \diff z \label{eq:Jeans1}\\
n \langle v_\phi^2 \rangle (R,z) &=& R \frac{\partial}{\partial R} \left( \frac{n \langle v_z^2 \rangle}{1-\beta_z} \right) + \frac{n \langle v_z^2 \rangle}{1 - \beta_z} + R n \frac{\partial \Phi}{\partial R},\label{eq:Jeans2}
\end{eqnarray}
where $n(\vect{x}) = \int f(\vect{x},\vect{v}) \Diff3 v$ is the number density of tracers and $\Phi(\vect{x})$ the galaxy's gravitational potential, generated by the mass density $\rho(\vect{x})$ via Poisson's equation.
\\The JAM modelling approach by \citet{Cap08} makes use of expressing the tracer density and the mass density as MGEs (see also \citet{1994A&A...285..723E}). The density of stellar tracers is assumed to be proportional to the observed and deprojected brightness distribution $\nu(R,z)$ in Eq. (\ref{eq:deprojMGE}). The mass density $\rho(R,z)$ consists of several sets of MGEs: One MGE, that is usually taken to be $\nu(R,z)$ multiplied by a constant stellar mass-to-light ratio $\Upsilon_*$, describes the distribution of stellar mass in the galaxy. (To mimick gradients of mass-to-light ratio, each Gaussian could be assigned its own $\Upsilon_{*,i}$.) To add a Navarro-Frenck-White (NFW) \citep{Navarro+1995c,NFW96} dark matter halo component, a MGE generated from a fit to a NFW profile can be added to the stellar component. [TO DO: Maybe remove detailed description of the mass MGE here, if it is described in detail somewhere else.] The gravitational potential is generated from the mass MGE by integrating the Poisson equation \citep{1994A&A...285..723E}. Equations \ref{eq:Jeans1} and \ref{eq:Jeans2} together with \ref{eq:bz} provide the velocity dispersion tensor $\langle v_i v_j \rangle = \langle v_i^2 \rangle$ (with $i,j \in \{ R,\phi, z\}$). It is then rotated by the inclination angle $i$ to the coordinate system of the observer ($(x',y')$ being the plane of the sky and $z'$ the line-of-sight, where $x'$ is alinged with the galaxy's major axis). Taking a light-weighted projection along the line-of-sight gives a model prediction for the line-of-sight velocity second moment, which is comparable to actual spectroscopic measurements of the second velocity moment. Details of the derivation using the MGE formalism are given in \citet{Cap08} and repeated in \citet{GlennEC}, we therefore just quote the result by \citet{Cap08} (their Equation 28) for the line-of-sight second velocity moment prediction from the Jeans equations,
\begin{eqnarray}
&&\left(I \langle v_\text{los}^2\rangle\right)(x',y')\\
&&= 4 \pi^{3/2} G \int_0^1 \sum_{k=1}^N \sum_{j=1}^M \nu_{0,k} \ q_j \ \rho_{0,j} \ u^2\nonumber\\
&& \times \frac{\sigma_k^2 \ q^{2}_k \left( \cos^2 i + \frac{\sin^2 i}{1-\beta_{z,k}}\right) + \mathscr{D} \  {x'}^{2} \sin^2 i}{(1-\mathscr{C}u^2) \sqrt{(\mathscr{A} + \mathscr{B} \cos^2 i) \left[1-(1-q^{2}_j) u^2 \right]}}\label{eq:explicitLOSvelCap}\\
&& \times \exp\left\{- \mathscr{A} \left[{x'}^{2} + \frac{(\mathscr{A}+\mathscr{B}) \ {y'}^{2}}{\mathscr{A}+\mathscr{B}\cos^2 i}\right] \right\} \diff u, \\
\end{eqnarray}
with $N$ Gaussians describing the tracer distribution and $M$ Gaussians describing the mass distribution, $\rho_{0,j}$ being the $j$-th mass density Gaussian evaluated at $(R=0,z=0)$ and
\begin{eqnarray*}
\mathscr{A} &=& \frac 12 \left(\frac{u^2}{\sigma_j^2} + \frac{1}{\sigma_k^2} \right)\nonumber\\
\mathscr{B} &=& \frac 12 \left\{\frac{1-q^{2}_k}{\sigma_k^2 q^{2}_k} + \frac{(1-q^{2}_j)u^4}{\sigma_j^2 \left[1-(1-{q}_j^{2})u^2 \right]} \right\}\nonumber\\
\mathscr{C} &=& 1- q^{2}_j - \frac{\sigma_k^2 q^{2}_k}{\sigma^2_j}\nonumber\\
\mathscr{D} &=& 1 - \frac{q^{2}_k}{1-\beta_{z,k}} - \left[ \left(1-\frac{1}{1-\beta_{k,z}}\right)\mathscr{C} + \frac{1-q^{2}_j}{1-\beta_{z,k}}\right] u^2.\nonumber
\end{eqnarray*}
The JAM modelling code by \citet{Cap08} evaluates numerically Equation \ref{eq:explicitLOSvelCap} for a given luminous tracer and mass distribution MGE and a given inclination.

[TO DO: make primed and unprimed coordinate systems clear.]

### 2015-09-03 docs/model_MGE_v1.01.tex
Revistied most of the lensing section.
Finished section on JAM theory.

\subsection{Multi-Gaussian Expansion Formalism}

@@

$I(R',\theta')$ describes the intrinsic, to 2D projected light distribution or surface density of the galaxy. Under the assumption that the galaxy is oblate and axisymmetric, and given the inclination angle $i$ of the galaxy with respect to the observer, MGEs allow an analytic deprojection of the 2D MGE to get a 3D light distribution or density $\nu(R,z)$ for the galaxy,
\begin{equation}
\nu(R,z) = \sum_i \nu_{0,i} \exp \left[-\frac{1}{2\sigma_i}\left(R^2 + \frac{z^2}{q_i^2} \right) \right]. \label{eq:deprojMGE}
\end{equation}
The flattening of each axisymmetric 3D Gaussian $q$ and its central density $\nu_{0,i}$ follow from the observed 2D axis ratio $q'_i$ and surface density $I_{0,i}$ as
\begin{eqnarray*}
q_i^2 &=& \frac{q_i'^2 - \cos^2 i}{\sin^2 i}\\
\nu_{0,i} &=& \frac{q_i' I_{0,i}}{q_i \sqrt{2 \pi \sigma_i^2}}.
\end{eqnarray*}

>>>>

\subsection{Multi-Gaussian Expansion Formalism} \label{sec:MGE_theo}

@@

$I(R',\theta')$ describes the intrinsic, to 2D projected light distribution or surface density of the galaxy. Under the assumption that the galaxy is oblate and axisymmetric, and given the inclination angle $i$ of the galaxy with respect to the observer, MGEs allow an analytic deprojection of the 2D MGE to get a 3D light distribution or density $\nu(R,z)$ for the galaxy,
\begin{equation}
\nu(R,z) = \sum_i \nu_{0,i} \exp \left[-\frac{1}{2\sigma_i}\left(R^2 + \frac{z^2}{q_i^2} \right) \right]. \label{eq:deprojMGE}
\end{equation}
The flattening of each axisymmetric 3D Gaussian $q_i$ and its central density $\nu_{0,i}$ follow from the observed 2D axis ratio $q'_i$ and surface density $I_{0,i}$ as
\begin{eqnarray*}
q_i^2 &=& \frac{q_i'^2 - \cos^2 i}{\sin^2 i}\\
\nu_{0,i} &=& \frac{q_i' I_{0,i}}{q_i \sqrt{2 \pi \sigma_i^2}}.
\end{eqnarray*}
